Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-19 paper.tex
Clarify n = 6 in depth benchmark

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=5,
        ymin=8, ymax=2^10,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        ymajorgrids=true,
        grid style=dashed
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Without precompilation}%
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        xmin=1, xmax=5,
        ymin=8, ymax=1024,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        %yticklabels={},
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{With precompilation}%
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{Results of the depth benchmark with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:erase}
\end{figure}

\paragraph{Depth Benchmark.}
Our final benchmark concerns situations where many rules match almost, but not quite.
We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
\jcom{What is $n$?}
Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.

This setup is challenging for the incremental approach because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$, big constant factors cause the incremental approach to remain slower for $k$ up to 5.

>>>>

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        ylabel={Time in ms},
        xmin=1, xmax=5,
        ymin=8, ymax=2^10,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        ymajorgrids=true,
        grid style=dashed
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 35.466292) (2, 40.918638) (3, 46.936125) (4, 54.933124) (5, 61.269861)
          };
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 135.967736) (2, 127.324444) (3, 136.891250) (4, 118.720444) (5, 116.293222)
          };
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 107.978069) (2, 115.900527) (3, 122.454958) (4, 137.732764) (5, 144.037681)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 619.932611) (2, 601.456805) (3, 600.926110) (4, 589.144389) (5, 595.091180)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Without precompilation}%
    \label{fig:erase-noprecomp}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Depth},
        xmin=1, xmax=5,
        ymin=8, ymax=1024,
        xtick={1,2,3,4,5},
        ymode=log,
        log basis y=2,
        %yticklabels={},
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 13.728805) (2, 17.909194) (3, 23.324666) (4, 28.030819) (5, 33.279611)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 13.841305) (2, 12.801791) (3, 13.358389) (4, 13.200611) (5, 13.251902)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          style=densely dashed,
          mark options={style={solid}}
          ]
          coordinates {
            (1, 81.546486) (2, 85.345444) (3, 90.532347) (4, 98.247764) (5, 105.423875)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.539097) (2, 24.432485) (3, 25.629666) (4, 23.935250) (5, 23.911139)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{With precompilation}%
    \label{fig:erase-precomp}
  \end{subfigure}
  \caption{Results of the depth benchmark for $n = 6$ with precompilation disabled (\subref{fig:erase-noprecomp}) or enabled (\subref{fig:erase-precomp}) and with $a = 0$ (\protect\showpgfsquare) or $a = 100$ (\protect\showpgfcircle).}%
  \label{fig:erase}
\end{figure}

\paragraph{Depth Benchmark.}
Our final benchmark concerns situations where many rules match almost, but not quite.
We use 100 copies of a rule $r : \All{x}{P_{1}~x → \dots → P_{n}~x → A}$, where the predicates $P_{i}$ are all distinct and $A$ is an arbitrary proposition.
Given a \emph{depth} $k ≤ n$, the initial goal contains hypotheses of types $P_{i}~a$ for each $i ≠ n - k$ with $1 ≤ i ≤ n$, as well as one hypothesis $P_{n-k}~(a + 1)$ (with $a = 0$ or $a = 100$).
In other words, the hypotheses match all premises with compatible substitutions, except the $k^{\text{th}}$ premise from the back.
Since the naive algorithm processes the premises back-to-front, $k$ determines how many premises the algorithm matches before it notices that the rule cannot be applied.

This setup is challenging for the incremental algorithm because (1) no rule is ever applied, so incrementality does not help; and (2) the incremental algorithm expends a non-negligible amount of effort for each of the $n$ hypotheses, whereas the naive algorithm can, at least for small $k$, decide fairly quickly that the rule is not applicable.
And indeed, Fig.~\ref{fig:erase-noprecomp} shows that while the naive approach slows down for greater $k$, big constant factors cause the incremental approach to remain slower for $k$ up to 5.
